Concurrent Programming in Harmony

### 2021-12-21 book/paper.tex
Updating book to use preproc Also renamed bag.count to bag.multiplicity to avoid clashes with variables called count

\index{choose operator}%
But programs can usually have more than one execution and produce multiple
different outputs as a result.  This is usually as a result of different
inputs, but Harmony programs do not have inputs. Instead,
\autoref{fig:choice} demonstrates
\emph{nondetermistic choice} in Harmony programs.  In this case, the
program chooses to print either ``hello'' or ``world''.  The corresponding
DFA captures both possibilities.
You can think of the \texttt{choose} operator as enumerating all possible
inputs to the program.

@@

\autoref{fig:triangle} shows another example of a Harmony program.
The example is a sequential program
and has a method \texttt{triangle} that takes
an integer number as argument.  Each method has a variable called
<{result}> that eventually contains the result of the
method (there is no \textbf{return} statement in Harmony).  The method
also has a bound variable called <{n}> containing the value of the
argument.  The <{{ x..y }}> notation generates a set containing the numbers
from~<{x}> to~<{y}> (inclusive).
(Harmony does not have iterators and in particular does
not have a \texttt{range} operator like Python.)
The last two lines in the program are
the most interesting.
The first assigns to <{x}> some unspecified value in the range \texttt{0..N}
and the second verifies that <{triangle(x)}> equals $x(x+1)/2$.

Running this Harmony program will try all possible executions, which
includes all possible values for $x$.  Try it out (here \texttt{\$}
represents a shell prompt):

@@

In Harmony, constants have a default specified value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for \texttt{N = 100}, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states 103
103 components, 0 bad states
No issues
$
\end{verbatim}
\end{code}

\section*{Exercises}
\begin{problems}
\item See what happens if, instead of initializing <{result}> to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:harmonymachine}.)
\item Write a Harmony program that computes squares by repeated adding.  So, the program
should compute the square of $x$ by adding $x$ to an initial value of 0 $x$ times.
\end{problems}

@@

Concurrent programming, aka multithreaded programming, involves multiple
threads
\index{thread}%
running in parallel while sharing variables.
\autoref{fig:progs} shows two programs.  Program (a) is sequential.
It sets <{shared}> to <{True}>, asserts that
<{shared}> = <{True}>, and finally sets
<{shared}> to <{False}>.
If you run the program through Harmony, it will not find any problems
because there is only one execution possible and 1) in that execution
the assertion does not fail and 2) the execution terminates.
Program (b) is concurrent---it executes methods \texttt{f()} and
\texttt{g()} in parallel.
If method \texttt{g()} runs and completes before \texttt{f()}, then
the assertion in \texttt{f()} will fail when \texttt{f()} runs.
Harmony will find this problem.
This problem is an example of non-determinism: methods \texttt{f()}
and \texttt{g()} can run in either order.

@@

\begin{problems}
\item Harmony programs can usually be easily translated into
Python by hand.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
If you're using Windows, the following batch script does the trick:
\begin{code}
\begin{verbatim}
FOR /L %%i IN (1, 1, 1000) DO python Up.py
PAUSE
\end{verbatim}
\end{code}
How many times does the assertion fail (if any)?
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment <{count}> \texttt{N} times.  Run \autoref{fig:incmany}
10 times (using Python).
Report how many times the assertion fails and what the value of <{count}>
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
(Try powers of 10 for \texttt{N}.)
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
(Also, you do not get to use the \texttt{atomically} keyword
or a \emph{lock}, yet.)
\end{problems}

@@

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
The Harmony compiler places the bytecode for file $x$.\texttt{htm} in
file $x$.\texttt{hvm}.
The model checker, called \emph{Charm}, places its output in a file
called $x$.\texttt{hco} and a stylized version of the same output
in an HTML file called $x$.\texttt{htm}.

@@

Using a lock is a good thought, but how does one implement one?
\autoref{fig:naivelock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by \textit{lockTaken} being <{False}>.
To enter the critical section, a thread waits until \textit{lockTaken} is <{False}>
and then sets it to <{True}> to indicate that the lock has been taken.
The thread then executes the critical section.  Finally the thread
releases the lock by setting \textit{lockTaken} back to <{False}>.

Unfortunately, if we run the program through Harmony,
we find that the assertion fails.
\autoref{fig:naiveLockhtml} shows the Harmony output.
Thread~0 finds that the lock is available, but just before it stores
<{True}> in \textit{lockTaken} in instruction~9, thread~1 gets
to run.
(Recall that you can hover your mouse over a machine instruction in order
to see what it does.)
Because \textit{lockTaken} is still <{False}>, it too
believes it can acquire the lock, and stores <{True}> in
\textit{lockTaken} and moves on to the critical section.
Finally, thread~0 moves on, also stores <{True}>
into \textit{lockTaken} and also moves
into the critical section.  Thread~0 is the one that detects the
problem.
The \textit{lockTaken} variable
suffers from the same sort of race condition as the <{count}> variable
in \autoref{fig:inc}: testing and setting the lock
consists of several instructions.  It is thus possible
for both threads to believe the lock is available and to obtain the lock
at the same time.

@@

To see why, first note the following invariant: if thread <{i}> is in the
critical section, then \textit{flags}[<{i}>] = <{True}>.
Without loss of generality,
suppose that thread~0 sets \textit{flags}[0] at time $t_0$.
Thread 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that \textit{flags}[1] = <{False}>.
Because of the invariant, \textit{flags}[1] = <{False}> implies that
thread~1 is not in the critical section at time $t_1$.
Let $t_2$ be the time at which thread~0 sets \textit{flags}[0]
to <{False}>.  Thread~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that thread~1 cannot enter the critical section
between $t_1$ and $t_2$, because \textit{flags}[1] = <{False}> at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set \textit{flags}[1] to <{True}> and
then wait until \textit{flags}[0] = <{False}>.  But that does not happen
until time $t_2$.

@@

The final na\"{\i}ve solution that we propose
is based on a variable called \textit{turn}.
Each thread politely lets the other thread have a turn first.
When \textit{turn} = <{i}>, thread~<{i}> can
enter the critical section, while thread $1-i$ has to wait.
An invariant of this solution is that while thread~<{i}> is in the critical
section, $\mathit{turn} = i$.
Since \textit{turn} cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that the thread that has been waiting the
longest to enter the critical section can go next.

@@

\glsadd{sequential consistency}%
In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:naiveflags} and \autoref{fig:naiveturn}, and is presented
in \autoref{fig:peterson}.
(The first line specifies that the \textit{flags} and \textit{turn}
variables should have \emph{sequential consistency}---it prevents
Harmony from complaining about data races involving these variables,
explained in \autoref{ch:spinlock}.)

@@

The HVM starts in an initial state in which there is only
one thread and its program counter is~0.  A \emph{trace}
\index{trace}%
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}%
in Harmony.
% \footnote{Actually, there is a third type called
% an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
One is when
there is more than one thread that can make a step.  The other is
when a thread executes a \texttt{choose} operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}%
if it is either the initial state
or it can be reached from the initial state through a finite trace.
A state is final
when there are no threads left to make state changes.

It is often useful to classify states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of classes of states.
\autoref{fig:states} depicts a Venn diagram of various classes of states
and a trace.
One way to classify states is to define a predicate over states.
All states in which $x$ = 1, or all states where
there are two or more threads executing, are examples of such predicates.
For our purposes, it is useful to define a predicate that says that at
most one thread is in the critical section.  We shall call such states
\emph{exclusive}.

@@

One candidate for such a predicate is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, exclusivity is not an \emph{inductive} invariant.
To see why, consider the following state $s$: let thread~0 be at label \texttt{cs}
and thread~1 be at the start of the \textbf{await} statement.
Also, in state $s$, $\mathit{turn} = 1$.  Now let
thread~1 make a step.  Because $\mathit{turn} = 1$,
thread~1 will stop waiting and also enter the critical
section, entering a state that is not exclusive.
So, exclusivity is an invariant (holds over every reachable state, as demonstrated
by Harmony),
but not an inductive invariant.
It will turn out that $s$ is not reachable.

@@

Instead, we will use the following property: $\mathcal{G}(i) =
\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@\mathtt{gate}$.
That is, if thread <{i}> is at the critical section, then
either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 8.
\autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
The label \texttt{gate} refers to the step that sets \textit{turn} to $1-i$.
You can run \autoref{fig:petersonproof} to determine
that $\mathcal{G}(i)$ is an invariant for $i = 0, 1$.
Moreover, if $\mathcal{F}(i)$ and $\mathcal{G}(i)$ both hold for $i = 0, 1$,
then mutual exclusion holds.  We can show this using proof by
contradiction.  Suppose mutual exclusion is violated and thus both threads are in
the critical section.  By $\mathcal{F}$ it must be the case that both
\textit{flags} are true.  By $\mathcal{G}$ and the fact that neither thread
is at label \texttt{gate}, we know that both $C(0)$ and $C(1)$ must hold.
This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
the desired contradiction.

@@

First consider the case that thread~0 is at label \texttt{cs}.  If thread~0
were to take a step, then in the next state thread~0 would be no longer
at that label and $\mathcal{G}(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by thread~1.
%
From $\mathcal{G}$ we know that one of the following three cases must hold before
thread~1 takes a step:
\begin{enumerate}
\item \textit{flags}[1] = <{False}>;
\item \textit{turn} = 0;
\item thread~1 is at label \texttt{gate}.
\end{enumerate}

Let us consider each of these cases.
We have to show that if thread~1 takes a step, then one of those
cases must hold after the step.
In the first case, if thread~1 takes a step, there are two possibilities:
either $flags[1]$ will still be <{False}> (in which case the first case
continues to hold), or $flags[1]$ will be <{True}>
and thread~1 will be at label \texttt{gate} (in which case the third case
will hold).
We know that thread~1 never sets \textit{turn} to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if thread~1 is at label \texttt{gate} before the step, then after
the step \textit{turn} will equal 0, and therefore the second case will hold
after the step.

@@

    thread X:                          thread Y:
        X0: flagX = True                   Y0: flagY = True
        X1: if not flagY goto X4           Y1: if not flagX goto Y4
        X2: flagX = False                  Y2: flagY = False
        X3: goto X0                        Y3: goto Y0
        X4: ...critical section...         Y4: ...critical section...
        X5: flagX = False                  Y5: flagY = False
\end{verbatim}
\end{code}
\item \label{ex:reorder}
A colleague of the author asked if the first two assignments in
Peterson's algorithm (setting \textit{flags}[\textit{self}])
to <{True}> and \textit{turn} to $1 - \mathit{self}$) can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.
See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in \autoref{fig:peterson} after reversing
the two assignments and describe in English what happens.
\item Bonus question:
Can you generalize Peterson's algorithm to more than two threads?
\item Bonus question:
Implement
\href{https://en.wikipedia.org/wiki/Dekker%27s_algorithm}{Dekker's Algorithm},
\href{https://en.wikipedia.org/wiki/Eisenberg_%26_McGuire_algorithm}{Eisenstein and McGuire's Algorithm},
\href{https://en.wikipedia.org/wiki/Szymanski%27s_algorithm}{Szymański's Algorithm}, or the
\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Lamport's Bakery Algorithm}.
Note that the last one uses unbounded state, so you should modify the threads so they
only try to enter the critical section a bounded number of times.
\end{problems}

@@

A method \texttt{m} with argument $a$ is invoked in its
most basic form as follows (assigning the result to~\texttt{r}).
\begin{code}
<{r = m a}>
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m}($a$), the argument is ($a$), which is the same
as $a$.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m}($a$, $b$), the argument is ($a$, $b$),
the tuple consisting of values $a$ and $b$.

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see \autoref{ch:harmonymachine}).
Both dictionaries and methods map Harmony values to Harmony values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and $x$ is an arbitrary Harmony value, then
\texttt{f} $x$, \texttt{f}($x$), and \texttt{f}[$x$] are all
the same expression in Harmony.

<{:const:P_enter}>
<{:const:P_exit}>
<{:const:P_mutex}>

@@

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If $x$ is a shared variable, then the expression ?$x$ is the address of $x$.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If $p$ is a pointer to a shared variable, then the
expression \texttt{!}$p$ is the value of the shared variable.
In particular, \texttt{!?}$x$ = $x$.
This is similar to how C pointers work (\texttt{*\string&}$x$ = $x$).

Often, pointers point to dictionaries, and so if $p$ is such a pointer,
then $(!p).\mathtt{field}$ would evaluate to the specified field in the dictionary.
Note that the parentheses in this expression
are needed, as \texttt{!}\textit{p}\texttt{.field} would wrongly evaluate
\texttt{!}(\textit{p}\texttt{.field}).
$(!p).\mathtt{field}$ is such a common expression that, like C, Harmony supports the
shorthand $p-$$>$\texttt{field}, which greatly improves readability.

\autoref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's shared memory state:
a turn variable and two flags.
Both methods \texttt{P\_enter} and \texttt{P\_exit} take two arguments:
a pointer to a mutex and the thread identifier (0 or 1).
$\mathit{pm}$$\rightarrow$$\mathtt{turn}$ is the value of the \texttt{.turn} key
in the dictionary that \textit{pm} points to.

You can put the first three methods in its own Harmony source file
and include it using the Harmony \texttt{import} statement.
\index{import statement}%
\index{module}%
This would make the code usable by multiple applications.

Finally, methods can have local variables.
Different from Python, Harmony local variables must be
explicitly bound within a scope.
Method variables are either mutable or immutable.
The arguments to a method and the bound variable (or
variables) within a \texttt{for} statement are immutable;
the variable <{result}> is mutable.
Using the <{var}> statement, new mutable local variables can be
declared.
For example, <{var}> $x = 3$ declares a new mutable local variable
$x$.
The \texttt{let} statement allows declaring new immutable local variables.
For example:
\texttt{let} $x = 3$: $y$ $+$$=$ $x$ adds 3 to the global variable
$y$.
See \autoref{sec:vars} for more information.

@@

This causes the context of the thread to be saved in $s$ and
the thread to be no longer running.
Another thread can revive the thread using the \textbf{go}
\index{go statement}%
statement:

>>>>

\index{choose operator}%
But programs can usually have more than one execution and produce multiple
different outputs as a result.  This is usually as a result of different
inputs, but Harmony programs do not have inputs. Instead,
\autoref{fig:choice} demonstrates
\emph{nondetermistic choice} in Harmony programs.  In this case, the
program chooses to print either ``hello'' or ``world''.  The corresponding
DFA captures both possibilities.
You can think of the <{choose}> operator as enumerating all possible
inputs to the program.

@@

\autoref{fig:triangle} shows another example of a Harmony program.
The example is a sequential program
and has a method <{triangle}> that takes
an integer number as argument.  Each method has a variable called
<{result}> that eventually contains the result of the
method (there is no \textbf{return} statement in Harmony).  The method
also has a bound variable called <{n}> containing the value of the
argument.  The <{{ x..y }}> notation generates a set containing the numbers
from~<{x}> to~<{y}> (inclusive).
(Harmony does not have iterators and in particular does
not have a \texttt{range} operator like Python.)
The last two lines in the program are
the most interesting.
The first assigns to <{x}> some unspecified value in the range \texttt{0..N}
and the second verifies that <{triangle(x)}> equals $x(x+1)/2$.

Running this Harmony program will try all possible executions, which
includes all possible values for <{x}>.  Try it out (here \texttt{\$}
represents a shell prompt):

@@

In Harmony, constants have a default specified value,
but those can be overridden on the command
line using the \texttt{-c} option.
\index{constant}%
For example, if you want to test the code for <{N = 100}>, run:
\begin{code}
\begin{verbatim}
$ harmony -c N=100 triangle.hny
#states 103
103 components, 0 bad states
No issues
$
\end{verbatim}
\end{code}

\section*{Exercises}
\begin{problems}
\item See what happens if, instead of initializing <{result}> to 0,
you initialize it to 1.  (You do not need to understand the error report at this time.  They will be explained in more detail in \autoref{ch:harmonymachine}.)
\item Write a Harmony program that computes squares by repeated adding.  So, the program
should compute the square of <{x}> by adding <{x}> to an initial value of 0 <{x}> times.
\end{problems}

@@

Concurrent programming, aka multithreaded programming, involves multiple
threads
\index{thread}%
running in parallel while sharing variables.
\autoref{fig:progs} shows two programs.  Program (a) is sequential.
It sets <{shared}> to <{True}>, asserts that
<{shared}> = <{True}>, and finally sets
<{shared}> to <{False}>.
If you run the program through Harmony, it will not find any problems
because there is only one execution possible and 1) in that execution
the assertion does not fail and 2) the execution terminates.
Program (b) is concurrent---it executes methods <{f()}> and
<{g()}> in parallel.
If method <{g()}> runs and completes before <{f()}>, then
the assertion in <{f()}> will fail when <{f()}> runs.
Harmony will find this problem.
This problem is an example of non-determinism: methods <{f()}>
and <{g()}> can run in either order.

@@

\begin{problems}
\item Harmony programs can usually be easily translated into
Python by hand.  For example,
\autoref{fig:incpy} is a Python version of \autoref{fig:inc}.
\begin{enumerate}
\item Run \autoref{fig:incpy} using Python.  Does the assertion fail?
\item Using a script, run \autoref{fig:incpy} 1000 times.
For example, if you are using the bash shell (in Linux or Mac OS X, say), you can do the following:
\begin{code}
\begin{verbatim}
for i in {1..1000}
do
    python Up.py
done
\end{verbatim}
\end{code}
If you're using Windows, the following batch script does the trick:
\begin{code}
\begin{verbatim}
FOR /L %%i IN (1, 1, 1000) DO python Up.py
PAUSE
\end{verbatim}
\end{code}
How many times does the assertion fail (if any)?
\end{enumerate}
\item \autoref{fig:incmany} is a version of \autoref{fig:incpy} that has each
incrementer thread increment <{count}> \texttt{N} times.  Run \autoref{fig:incmany}
10 times (using Python).
Report how many times the assertion fails and what the value of <{count}>
was for each of the failed runs.
Also experiment with lower values of \texttt{N}.
How large does \texttt{N} need to be for assertions to fail?
(Try powers of 10 for \texttt{N}.)
\item Can you think of a fix to \autoref{fig:inc}?  Try one or two different fixes
and run them through Harmony.  Do not worry about having to come up with a correct fix at this
time---the important thing is to develop an understanding of concurrency.
(Also, you do not get to use the <{atomically}> keyword
or a \emph{lock}, yet.)
\end{problems}

@@

Harmony programs are compiled to Harmony \emph{bytecode}
\index{bytecode}%
(a list of machine instructions for a virtual machine),
which in turn is executed by the Harmony virtual machine (HVM).
\index{virtual machine}%
\index{Harmony Virtual Machine}%
\index{HVM}%
The Harmony compiler places the bytecode for file <{x}>.\texttt{htm} in
file <{x}>.\texttt{hvm}.
The model checker, called \emph{Charm}, places its output in a file
called <{x}>.\texttt{hco} and a stylized version of the same output
in an HTML file called <{x}>.\texttt{htm}.

@@

Using a lock is a good thought, but how does one implement one?
\autoref{fig:naivelock} presents a mutual exclusion attempt based on a
na\"{\i}ve (and, as it turns out, broken) implementation of a lock.
Initially the lock is not owned, indicated by <{lockTaken}> being <{False}>.
To enter the critical section, a thread waits until <{lockTaken}> is <{False}>
and then sets it to <{True}> to indicate that the lock has been taken.
The thread then executes the critical section.  Finally the thread
releases the lock by setting <{lockTaken}> back to <{False}>.

Unfortunately, if we run the program through Harmony,
we find that the assertion fails.
\autoref{fig:naiveLockhtml} shows the Harmony output.
Thread~0 finds that the lock is available, but just before it stores
<{True}> in <{lockTaken}> in instruction~9, thread~1 gets
to run.
(Recall that you can hover your mouse over a machine instruction in order
to see what it does.)
Because <{lockTaken}> is still <{False}>, it too
believes it can acquire the lock, and stores <{True}> in
<{lockTaken}> and moves on to the critical section.
Finally, thread~0 moves on, also stores <{True}>
into <{lockTaken}> and also moves
into the critical section.  Thread~0 is the one that detects the
problem.
The <{lockTaken}> variable
suffers from the same sort of race condition as the <{count}> variable
in \autoref{fig:inc}: testing and setting the lock
consists of several instructions.  It is thus possible
for both threads to believe the lock is available and to obtain the lock
at the same time.

@@

To see why, first note the following invariant: if thread <{i}> is in the
critical section, then <{flags[i] = True}>.
Without loss of generality,
suppose that thread~0 sets <{flags[0]}> at time $t_0$.
Thread 0 can only reach the critical section if at some time $t_1$,
$t_1 > t_0$, it finds that <{flags[1] = False}>.
Because of the invariant, <{flags[1] = False}> implies that
thread~1 is not in the critical section at time $t_1$.
Let $t_2$ be the time at which thread~0 sets <{flags[0]}>
to <{False}>.  Thread~0 is in the critical section sometime
between $t_1$ and $t_2$.
It is easy to see that thread~1 cannot enter the critical section
between $t_1$ and $t_2$, because <{flags[1] = False}> at
time $t_1$.  To reach the critical section between $t_1$ and $t_2$,
it would first have to set <{flags[1]}> to <{True}> and
then wait until <{flags[0] = False}>.  But that does not happen
until time $t_2$.

@@

The final na\"{\i}ve solution that we propose
is based on a variable called <{turn}>.
Each thread politely lets the other thread have a turn first.
When <{turn}> = <{i}>, thread~<{i}> can
enter the critical section, while thread $1-i$ has to wait.
An invariant of this solution is that while thread~<{i}> is in the critical
section, $\mathit{turn} = i$.
Since <{turn}> cannot be 0 and 1 at
the same time, mutual exclusion is satisfied.
The solution also has the nice property that the thread that has been waiting the
longest to enter the critical section can go next.

@@

\glsadd{sequential consistency}%
In 1981, Gary L.~Peterson came up with a beautiful solution to the mutual exclusion
problem, now known as ``Peterson's Algorithm''~\cite{Peterson81}.
The algorithm is an amalgam of the (incorrect) algorithms in
\autoref{fig:naiveflags} and \autoref{fig:naiveturn}, and is presented
in \autoref{fig:peterson}.
(The first line specifies that the \textit{flags} and <{turn}>
variables should have \emph{sequential consistency}---it prevents
Harmony from complaining about data races involving these variables,
explained in \autoref{ch:spinlock}.)

@@

The HVM starts in an initial state in which there is only
one thread and its program counter is~0.  A \emph{trace}
\index{trace}%
is a sequence of steps starting from the initial state.
When making a step, there are two sources of non-determinism
\index{non-determinism}%
in Harmony.
% \footnote{Actually, there is a third type called
% an \emph{interrupt}, discussed in \autoref{ch:interrupts}.}
One is when
there is more than one thread that can make a step.  The other is
when a thread executes a <{choose}> operation and there is
more than one choice.
Because there is non-determinism, there are multiple possible traces.
We call a state \emph{reachable}
\index{reachable state}%
if it is either the initial state
or it can be reached from the initial state through a finite trace.
A state is final
when there are no threads left to make state changes.

It is often useful to classify states.
\emph{Initial}, \emph{final}, and \emph{reachable}, and \emph{unreachable}
are all examples of classes of states.
\autoref{fig:states} depicts a Venn diagram of various classes of states
and a trace.
One way to classify states is to define a predicate over states.
All states in which <{x}> = 1, or all states where
there are two or more threads executing, are examples of such predicates.
For our purposes, it is useful to define a predicate that says that at
most one thread is in the critical section.  We shall call such states
\emph{exclusive}.

@@

One candidate for such a predicate is exclusivity itself.
After all, it certainly holds over the initial state.
And as Harmony has already determined, exclusivity is an invariant:
it holds over every reachable state.
Unfortunately, exclusivity is not an \emph{inductive} invariant.
To see why, consider the following state <{s}>: let thread~0 be at label \texttt{cs}
and thread~1 be at the start of the \textbf{await} statement.
Also, in state <{s}>, $\mathit{turn} = 1$.  Now let
thread~1 make a step.  Because $\mathit{turn} = 1$,
thread~1 will stop waiting and also enter the critical
section, entering a state that is not exclusive.
So, exclusivity is an invariant (holds over every reachable state, as demonstrated
by Harmony),
but not an inductive invariant.
It will turn out that <{s}> is not reachable.

@@

Instead, we will use the following property: $\mathcal{G}(i) =
\mathtt{thread}(i)@\mathtt{cs} \Rightarrow \mathcal{C}(i) \lor \mathtt{thread}(1-i)@\mathtt{gate}$.
That is, if thread <{i}> is at the critical section, then
either $\mathcal{C}(i)$ holds or thread $1-i$ is about to execute line 8.
\autoref{fig:petersonproof} formalizes $\mathcal{G}(i)$ in Harmony.
The label \texttt{gate} refers to the step that sets <{turn}> to $1-i$.
You can run \autoref{fig:petersonproof} to determine
that $\mathcal{G}(i)$ is an invariant for $i = 0, 1$.
Moreover, if $\mathcal{F}(i)$ and $\mathcal{G}(i)$ both hold for $i = 0, 1$,
then mutual exclusion holds.  We can show this using proof by
contradiction.  Suppose mutual exclusion is violated and thus both threads are in
the critical section.  By $\mathcal{F}$ it must be the case that both
\textit{flags} are true.  By $\mathcal{G}$ and the fact that neither thread
is at label \texttt{gate}, we know that both $C(0)$ and $C(1)$ must hold.
This then implies that $\mathit{turn} = 0 \land \mathit{turn} = 1$, providing
the desired contradiction.

@@

First consider the case that thread~0 is at label \texttt{cs}.  If thread~0
were to take a step, then in the next state thread~0 would be no longer
at that label and $\mathcal{G}(0)$ would hold trivially over the next state.
Therefore we only need to consider a step by thread~1.
%
From $\mathcal{G}$ we know that one of the following three cases must hold before
thread~1 takes a step:
\begin{enumerate}
\item \textit{flags}[1] = <{False}>;
\item <{turn}> = 0;
\item thread~1 is at label \texttt{gate}.
\end{enumerate}

Let us consider each of these cases.
We have to show that if thread~1 takes a step, then one of those
cases must hold after the step.
In the first case, if thread~1 takes a step, there are two possibilities:
either $flags[1]$ will still be <{False}> (in which case the first case
continues to hold), or $flags[1]$ will be <{True}>
and thread~1 will be at label \texttt{gate} (in which case the third case
will hold).
We know that thread~1 never sets <{turn}> to 1, so
if the second case holds before the step, it will also hold after the step.
Finally, if thread~1 is at label \texttt{gate} before the step, then after
the step <{turn}> will equal 0, and therefore the second case will hold
after the step.

@@

    thread X:                          thread Y:
        X0: flagX = True                   Y0: flagY = True
        X1: if not flagY goto X4           Y1: if not flagX goto Y4
        X2: flagX = False                  Y2: flagY = False
        X3: goto X0                        Y3: goto Y0
        X4: ...critical section...         Y4: ...critical section...
        X5: flagX = False                  Y5: flagY = False
\end{verbatim}
\end{code}
\item \label{ex:reorder}
A colleague of the author asked if the first two assignments in
Peterson's algorithm (setting <{flags[self]}>
to <{True}> and <{turn}> to <{1 - self}> can be reversed.
After all, they are different variables assigned independent values---in a
sequential program one could surely swap the two assignments.
See if you can figure out for yourself if the two assignments can be
reversed.  Then run the program in \autoref{fig:peterson} after reversing
the two assignments and describe in English what happens.
\item Bonus question:
Can you generalize Peterson's algorithm to more than two threads?
\item Bonus question:
Implement
\href{https://en.wikipedia.org/wiki/Dekker%27s_algorithm}{Dekker's Algorithm},
\href{https://en.wikipedia.org/wiki/Eisenberg_%26_McGuire_algorithm}{Eisenstein and McGuire's Algorithm},
\href{https://en.wikipedia.org/wiki/Szymanski%27s_algorithm}{Szymański's Algorithm}, or the
\href{https://en.wikipedia.org/wiki/Lamport%27s_bakery_algorithm}{Lamport's Bakery Algorithm}.
Note that the last one uses unbounded state, so you should modify the threads so they
only try to enter the critical section a bounded number of times.
\end{problems}

@@

A method <{m}> with argument <{a}> is invoked in its
most basic form as follows (assigning the result to~\texttt{r}).
\begin{code}
<{r = m a}>
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
<{m(a)}>, the argument is (<{a}>), which is the same
as <{a}>.
If you invoke <{m()}>, the argument is <{()}>,
which is the empty tuple.
If you invoke <{m(a, b)}>, the argument is <{(a, b)}>,
the tuple consisting of values <{a}> and <{b}>.

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see \autoref{ch:harmonymachine}).
Both dictionaries and methods map Harmony values to Harmony values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and <{x}> is an arbitrary Harmony value, then
<{f x}>, <{f(x)}>, and <{f[x]}> are all
the same expression in Harmony.

<{:const:P_enter}>
<{:const:P_exit}>
<{:const:P_mutex}>
<{:var:p}>

@@

Harmony is not an object-oriented language like Python is.  In Python
you can pass a reference to an object to a method, and that method
can then update the object.  In Harmony, it is also sometimes convenient
to have a method update a shared variable specified as an argument.
For this, as mentioned in \autoref{ch:harmonymachine},
each shared variable has an \emph{address}, itself a Harmony value.
\index{address}%
If <{x}> is a shared variable, then the expression <{?x}> is the address of <{x}>.
If a variable contains an address, we call that variable a \emph{pointer}.
\index{pointer}%
If <{p}> is a pointer to a shared variable, then the
expression \texttt{!}<{p}> is the value of the shared variable.
In particular, <{!?x = x}>.
This is similar to how C pointers work (\texttt{*\string&}<{x}> = <{x}>).

Often, pointers point to dictionaries, and so if <{p}> is such a pointer,
then <{(!p).field}> would evaluate to the specified field in the dictionary.
Note that the parentheses in this expression
are needed, as <{!p.field}> would wrongly evaluate
<{!(p.field)}>.
<{(!p).field}> is such a common expression that, like C, Harmony supports the
shorthand <{p->field}>, which greatly improves readability.

\autoref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
The name \textit{mutex} is often used to denote a variable or value
that is used for mutual exclusion.
\texttt{P\_mutex} is a method that returns a ``mutex,'' which, in this
case, is a dictionary that contains Peterson's Algorithm's shared memory state:
a turn variable and two flags.
Both methods <{P_enter}> and <{P_exit}> take two arguments:
a pointer to a mutex and the thread identifier (0 or 1).
<{pm->turn}> is the value of the <{.turn}> key
in the dictionary that <{pm}> points to.

You can put the first three methods in its own Harmony source file
and include it using the Harmony <{import}> statement.
\index{import statement}%
\index{module}%
This would make the code usable by multiple applications.

Finally, methods can have local variables.
Different from Python, Harmony local variables must be
explicitly bound within a scope.
Method variables are either mutable or immutable.
The arguments to a method and the bound variable (or
variables) within a <{for}> statement are immutable;
the variable <{result}> is mutable.
Using the <{var}> statement, new mutable local variables can be
declared.
For example, <{var}> $x = 3$ declares a new mutable local variable
<{x}>.
The <{let}> statement allows declaring new immutable local variables.
For example:
<{let}> $x = 3$: <{y}> $+$$=$ <{x}> adds 3 to the global variable
<{y}>.
See \autoref{sec:vars} for more information.

@@

This causes the context of the thread to be saved in <{s}> and
the thread to be no longer running.
Another thread can revive the thread using the \textbf{go}
\index{go statement}%
statement:
